---
title: "Seed aging"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Seed aging}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  out.width = "100%",
  dpi = 96
)
```

```{r setup}
library(pbtm)
```

## The model

As seeds age, germination first slows while viability stays nearly constant.
Viability then declines along a sigmoid curve. Seeds in a lot have a range of
potential lifetimes that follow a normal distribution (Ellis and Roberts,
1981). Bradford et al. (1993) used the increasing germination delays to predict
that decline with a threshold model:

$$\theta_A = (p_{max}(g) - p)\, t_g$$

where $p$ is the aging (storage) period, $p_{max}(g)$ is the maximum lifetime
of seed fraction `g`, normally distributed with median $p_{max}(50)$ and
standard deviation $\sigma$, and $\theta_A$ is the aging time constant.
Germination falls as the aging period approaches each seed's lifetime, so the
fitted model uses the upper tail of the distribution:

$$g = 1 - \Phi\left(\frac{p + \theta_A / t_g - p_{max}(50)}{\sigma}\right)$$

## Fitting

`aging_data` holds germination time courses of lettuce seeds after 0 to 6 days
of accelerated aging:

```{r}
plot_germ_data(aging_data, color = "AgingTime")
fit <- fit_aging(aging_data)
fit
autoplot(fit)
```

Half of the seeds would lose viability after
`r round(coef(fit)[["p_max50"]], 1)` days of accelerated aging, with a standard
deviation of `r round(coef(fit)[["sigma"]], 1)` days. The units of $\theta_A$
are aging time × germination time (here day·hours).

On the normalized axis, each observation is placed at the aging threshold
implied by its germination time, $p + \theta_A / t$. The line shows the
fraction of seeds whose lifetime exceeds that threshold:

```{r}
autoplot(fit, type = "normalized")
```

## Predicting the loss of viability

The fitted distribution predicts the viability curve beyond the aging periods
tested, which is the main practical use of the model. Viability after aging
period $p$ is the fraction of seeds with $p_{max} > p$:

```{r}
p <- coef(fit)
aging <- seq(0, 25, by = 5)
data.frame(
  AgingTime = aging,
  viability = pnorm(aging, p[["p_max50"]], p[["sigma"]], lower.tail = FALSE)
)
```

Accelerated-aging results must be calibrated against the temperature and
moisture conditions under which the seeds will actually be stored (Bradford and
Bello, 2023).

## References

Bradford, K.J., Bello, P. (2023) Understanding seed behavior: populations of
individuals. *Acta Hortic.* 1365: 1–16.
<https://doi.org/10.17660/ActaHortic.2023.1365.1>

Bradford, K.J., Tarquis, A.M., Duran, J.M. (1993) A population-based threshold
model describing the relationship between germination rates and seed
deterioration. *J. Exp. Bot.* 44: 1225–1234.
<https://doi.org/10.1093/jxb/44.7.1225>

Ellis, R.H., Roberts, E.H. (1981) The quantification of ageing and survival in
orthodox seeds. *Seed Sci. Technol.* 9: 373–409.
