---
title: "Hydrotime and hydrothermal time"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Hydrotime and hydrothermal time}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  out.width = "100%",
  dpi = 96
)
```

```{r setup}
library(pbtm)
```

## Hydrotime

Seeds germinate more slowly, and fewer of them germinate, as the water
potential $\psi$ of their surroundings decreases. The hydrotime model
(Gummerson, 1986; Bradford, 1990) states that the *hydrotime* to germination is
constant:

$$\theta_H = (\psi - \psi_b(g))\, t_g$$

where $t_g$ is the time to germination of fraction `g`. The base water
potential $\psi_b(g)$, below which a seed cannot germinate, varies normally
across the population with median $\psi_b(50)$ and standard deviation
$\sigma$. Solving for the germinated fraction gives the fitted model:

$$g = \Phi\left(\frac{\psi - \theta_H / t_g - \psi_b(50)}{\sigma}\right)$$

`hydrotime_data` has tomato seeds germinated at 20 °C and three water
potentials. Hydrotime data should come from a single temperature; filter other
temperatures out first.

```{r}
plot_germ_data(hydrotime_data, color = "GermWP")
fit_ht <- fit_hydrotime(hydrotime_data)
fit_ht
```

Half the seeds can germinate at water potentials down to
`r round(coef(fit_ht)[["psi_b50"]], 2)` MPa. The hydrotime constant is
`r round(coef(fit_ht)[["theta_h"]])` MPa·h.

```{r}
autoplot(fit_ht)
```

On a probit fraction axis the fitted curves become nearly straight. The
normalized plot places every observation at the base water potential implied
by its germination time, $\psi - \theta_H / t$. All water potentials then fall
on one line, the population's distribution of $\psi_b$:

```{r}
autoplot(fit_ht, y_scale = "probit")
autoplot(fit_ht, type = "normalized")
```

The median germination time at any water potential above $\psi_b(50)$ is
$t_{50} = \theta_H / (\psi - \psi_b(50))$:

```{r}
p <- coef(fit_ht)
wp <- c(0, -0.25, -0.4)
data.frame(GermWP = wp, predicted_t50 = p[["theta_h"]] / (wp - p[["psi_b50"]]))
germ_speed(hydrotime_data, 0.5, groups = "GermWP")
```

The predictions are close to the observed medians at -0.25 and -0.4 MPa. In
this example dataset, however, the seeds in water (0 MPa) germinated more
slowly than those at -0.25 MPa, contrary to the model. The normalized plot above
shows this: the 0 MPa points fall to the right of the line in its upper half,
and the treatments do not collapse perfectly onto it. Departures like this are
worth checking against the experimental records before drawing conclusions
from the parameters.

## Hydrothermal time

The hydrothermal time model combines the thermal and hydrotime models to
describe germination across both temperature and water potential
(Gummerson, 1986; Bradford, 1995):

$$\theta_{HT} = (\psi - \psi_b(g))\,(T - T_b)\, t_g$$

$$g = \Phi\left(\frac{\psi - \theta_{HT} / [(T - T_b)\, t_g] - \psi_b(50)}{\sigma}\right)$$

`hydrothermal_time_data` crosses three temperatures with three water potentials:

```{r}
fit_htt <- fit_hydrothermal_time(hydrothermal_time_data)
fit_htt
autoplot(fit_htt)
```

The normalized plot uses both estimated constants, $\theta_{HT}$ and $T_b$, to
put all nine treatments on a common base water potential scale:

```{r}
autoplot(fit_htt, type = "normalized")
```

Systematic departures from a single line, for example one temperature lying
consistently above the others, show where the assumption of a constant
$\psi_b$ distribution across temperatures breaks down. In practice $\psi_b(50)$
often shifts with temperature, particularly above the optimum.

## References

Bradford, K.J. (1990) A water relations analysis of seed germination rates.
*Plant Physiol.* 94: 840–849. <https://doi.org/10.1104/pp.94.2.840>

Bradford, K.J. (1995) Water relations in seed germination. In J Kigel, G
Galili, eds, *Seed Development and Germination*. Marcel Dekker, New York,
pp 351–396.

Gummerson, R.J. (1986) The effect of constant temperatures and osmotic
potentials on the germination of sugar beet. *J. Exp. Bot.* 37: 729–741.
