---
title: "Getting started with pbtm"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with pbtm}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  out.width = "100%",
  dpi = 96
)
```

```{r setup}
library(pbtm)
```

Population-based threshold (PBT) models describe how a population of seeds
responds to a factor such as temperature, water potential, a hormone, or aging.
Each seed has its own *threshold* for the factor, and thresholds vary normally
across the population. When the factor exceeds a seed's threshold, that seed
germinates at a rate proportional to the amount by which the threshold is
exceeded. Fitting a PBT model to germination time courses collected at several
factor levels estimates the threshold distribution (its median and standard
deviation) and a time constant for the response, from which germination under
any other condition can be predicted (Bradford and Bello, 2022).

pbtm fits these models with nonlinear least squares and plots the results. This
vignette covers data preparation and the tools shared by every model; each
model has its own vignette:

* `vignette("thermal-time")`: base temperature and thermal time.
* `vignette("hydrotime")`: hydrotime and hydrothermal time.
* `vignette("priming")`: hydropriming and hydrothermal priming.
* `vignette("aging")`: seed aging.
* `vignette("promoters-inhibitors")`: hormone dose responses.
* `vignette("subpopulations")`: seed lots made of several subpopulations.

## Preparing data

pbtm expects one row per observation of a germination time course, with the
cumulative elapsed time (`CumTime`) and the cumulative fraction of seeds
germinated by that time (`CumFraction`, from 0 to 1). Treatment columns repeat
the treatment's value on every row of its time course. `pbtm_columns` lists
the template columns, their valid ranges, and which models use them:

```{r}
pbtm_columns[, c("Column", "Description", "Min", "Max")]
```

`TrtID` identifies each time course and `TrtDesc` describes it. Every model
needs `CumTime` and `CumFraction` plus its own treatment columns, which
`pbtm_models()` lists:

```{r}
pbtm_models()[, c("model", "label", "factors", "params")]
```

The package includes an example dataset for each model (see `?pbtm_datasets`):

```{r}
head(germination_data)
```

If your data are raw counts, `calc_cum_frac()` converts cumulative counts of
germinated seeds into `CumFraction`. `validate_germ_data()` checks column types
and ranges, and optionally that a model's columns are present:

```{r}
validate_germ_data(hydrotime_data, "hydrotime")
```

Data with different column names can be used without renaming them by passing
a `cols` mapping from template names to your names to `validate_germ_data()`
or any fitting function:

```{r}
my_data <- dplyr::rename(thermal_time_data, temp = GermTemp, hours = CumTime)
fit <- fit_thermal_time(my_data, cols = c(GermTemp = "temp", CumTime = "hours"))
```

## Exploring germination time courses

`plot_germ_data()` plots the raw time courses:

```{r}
plot_germ_data(germination_data, color = "GermTemp")
```

`germ_speed()` interpolates the time for each treatment to reach given
germination fractions and the corresponding germination rate
(`GR = 1 / Time`). The 16% and 84% fractions are one standard deviation either
side of the median:

```{r}
germ_speed(germination_data, fractions = c(0.1, 0.16, 0.5, 0.84, 0.9))
```

To summarize over several treatments at once, name the columns to group by.
Time courses sharing those values are pooled with `rescale_cum_frac()`, which
adds up their germination increments and re-accumulates them into one curve.

Observations where germination did not increase since the previous time point
add no information about individual seeds' germination times but still pull on
the fitted curves. `clean_germ_data()` removes them, keeping the first time each
cumulative fraction was reached. Using cleaned data is recommended for most
analyses:

```{r}
nrow(promoter_data)
nrow(clean_germ_data(promoter_data))
```

## Fitting a model

Each model has a fitting function (`fit_thermal_time()`, `fit_hydrotime()`,
...), and `fit_pbtm()` fits any model by its id. All of them share the same
options:

* `max_frac`: the highest fraction the population can reach, for seed lots
  that do not germinate fully even under optimal conditions.
* `fixed`: parameter values to hold constant, e.g. `fixed = list(t_b = 5)`.
* `bounds`: custom `c(lower, start, upper)` bounds for any parameter (defaults
  are in `pbtm_models()$bounds`).
* `subpops`: fit a mixture of seed subpopulations (see
  `vignette("subpopulations")`).

```{r}
fit <- fit_hydrotime(hydrotime_data)
fit
```

The fit works with the usual model functions:

```{r}
coef(fit)
summary(fit)
predict(fit, newdata = data.frame(GermWP = -0.3, CumTime = c(100, 200, 300)))
```

If an estimate ends up on one of its bounds, or the fit does not converge,
pbtm warns you. This usually means the data do not constrain that parameter
well, and it can be held `fixed` or given wider `bounds`.

## Plotting fits and linearized scales

`autoplot()` (or `plot()`) draws the data with the fitted model:

```{r}
autoplot(fit)
```

The cumulative models are normal distributions of thresholds, so their S-shaped
curves become close to straight lines when the fraction axis uses the normal
quantile, or *probit*, scale. For thermal time, whose distribution is
log-normal in time, a log time axis combined with a probit fraction axis makes
the fitted curves exactly straight. Deviations from a straight line then point
to where the model fits poorly. Observations at 0% or 100% germination cannot
be drawn on a probit axis and are dropped with a message.

```{r}
autoplot(fit, x_scale = "log", y_scale = "probit")
```

The `"normalized"` plot goes a step further: it places each observation on the
model's threshold axis (here, the base water potential that seed must have had,
`GermWP - theta_h / CumTime`). All treatments then collapse onto the single
population distribution, a straight line on the probit scale, with the dashed
line marking the median:

```{r}
autoplot(fit, type = "normalized")
```

## References

Bradford, K.J., Bello, P. (2022) Applying population-based threshold models to
quantify and improve seed quality attributes. In J Buitink, O Leprince, eds,
*Advances in Seed Science and Technology for More Sustainable Crop Production*.
Burleigh Dodds Science Publishing, Cambridge, UK.
<https://doi.org/10.19103/AS.2022.0105.05>
