---
title: "Hydropriming and hydrothermal priming"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Hydropriming and hydrothermal priming}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  out.width = "100%",
  dpi = 96
)
```

```{r setup}
library(pbtm)
```

Priming hydrates seeds to just below the level needed to complete germination,
holds them there for a while, and dries them again (Bradford, 1986). Primed
seeds germinate faster and more uniformly. The advancement depends on the
priming water potential and duration, and, for hydrothermal priming, on the
priming temperature (Bradford and Haigh, 1994).

Unlike the other models, the priming models are fit to germination *rates*.
The data are germination time courses of seed samples primed under different
conditions and then germinated under standard conditions. The fitting
functions first compute the germination rate of each priming treatment at a
chosen fraction (`speed = 0.5` by default, i.e. $GR_{50}$, the inverse of the
time to 50% germination) with `germ_speed()`, and then fit a line to those
rates.

## Hydropriming

The hydropriming time accumulated during priming at water potential $\psi_P$
for duration $t_P$ is $(\psi_P - \psi_{min})\,t_P$, where $\psi_{min}$ is the
lowest water potential that still has a priming effect. Germination rates
increase linearly with it:

$$GR_{50} = GR_i + \frac{(\psi_P - \psi_{min})\,t_P}{\theta_{HP}}$$

where $GR_i$ is the rate of unprimed seeds and $\theta_{HP}$ is the
hydropriming time constant. pbtm estimates the `slope`, $1/\theta_{HP}$.

`hydropriming_data` includes unprimed controls and priming at three water
potentials for up to 148 hours:

```{r}
fit_hp <- fit_hydropriming(hydropriming_data)
fit_hp
```

The germination rates the model was fit to are in `fit_hp$data`:

```{r}
head(fit_hp$data)
```

The rates lie on a line against hydropriming time, which requires a
`psi_min` estimate to compute:

```{r}
autoplot(fit_hp)
```

Log axes on both scales (`x_scale = "log", y_scale = "log"`) are also
available. Because of the $GR_i$ intercept, the fitted line curves on a log-log
plot.

### Designing a priming treatment

Any combination of water potential and duration that accumulates the same
hydropriming time has the same effect. Rearranging the model gives the duration
needed at a chosen water potential to reach a target rate:

```{r}
p <- coef(fit_hp)
target_gr <- 0.02
psi <- c(-0.5, -1)
data.frame(
  PrimingWP = psi,
  duration = (target_gr - p[["gr_i"]]) / (p[["slope"]] * (psi - p[["psi_min"]]))
)
```

`predict()` gives the expected rate for any proposed treatment:

```{r}
predict(fit_hp, newdata = data.frame(PrimingWP = -0.5, PrimingDuration = c(24, 48, 96)))
```

## Hydrothermal priming

Hydrothermal priming time also counts the temperature above a minimum
temperature $T_{min}$ for a priming effect:

$$GR_{50} = GR_i + \frac{(\psi_P - \psi_{min})(T_P - T_{min})\,t_P}{\theta_{HTP}}$$

```{r}
fit_htp <- fit_hydrothermal_priming(hydrothermal_priming_data)
fit_htp
autoplot(fit_htp)
```

Warmer or wetter priming shortens the duration needed, and cooler or drier
priming lengthens it. The same effect can therefore be reached with whatever
combination of temperature, water potential, and duration is most convenient.

## Rates at other fractions

To base the model on a different part of the germination time course, change
`speed`, for example to the time to 25% germination:

```{r}
fit_hydropriming(hydropriming_data, speed = 0.25)
```

If you already have a table of germination rates, pass it with a `GR` column
(and no `CumFraction` column) and it is used directly.

## References

Bradford, K.J. (1986) Manipulation of seed water relations via osmotic priming
to improve germination under stress conditions. *HortScience* 21: 1105–1112.

Bradford, K.J., Haigh, A.M. (1994) Relationship between accumulated
hydrothermal time during seed priming and subsequent seed germination rates.
*Seed Sci. Res.* 4: 1–10.
