---
title: "Promoters and inhibitors"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Promoters and inhibitors}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  out.width = "100%",
  dpi = 96
)
```

```{r setup}
library(pbtm)
```

Chemical factors can speed up or slow down germination. Hormones are the
classic example: gibberellin (GA) breaks dormancy and promotes germination,
while abscisic acid (ABA) induces dormancy and inhibits it. The same threshold
logic applies: each seed has a threshold dose, and the response rate depends on
how far the applied dose is from it (Ni and Bradford, 1992, 1993).

Hormone dose responses are usually linear on a log scale, so both models take
`dose_transform = "log10"`. The threshold parameters (`p_b50`, `i_b50`, and
`sigma`) are then in log10 dose units.

## Promoters

$$\theta_P = (\log[P] - \log[P_b(g)])\, t_g$$

$$g = \Phi\left(\frac{\log[P] - \theta_P / t_g - \log P_b(50)}{\sigma}\right)$$

`promoter_data` has tomato seeds germinated with three GA concentrations:

```{r}
plot_germ_data(promoter_data, color = "GermPromoterDosage")
fit_p <- fit_promoter(promoter_data, dose_transform = "log10")
fit_p
autoplot(fit_p)
```

The median base GA dose is $10^{`r round(coef(fit_p)[["p_b50"]], 3)`}$ =
`r signif(10^coef(fit_p)[["p_b50"]], 3)` dose units.

The choice of scale matters. On the linear dose scale the fit is much worse,
and the time constant runs into its upper bound, which pbtm reports as a
warning:

```{r}
fit_linear <- fit_promoter(promoter_data)
c(log10 = fit_p$stats$pseudo_r2, linear = fit_linear$stats$pseudo_r2)
```

The normalized plot uses the threshold axis $\log[P] - \theta_P / t$:

```{r}
autoplot(fit_p, type = "normalized")
```

## Inhibitors

Inhibitors act in the opposite direction: germination slows and declines as the
dose approaches each seed's inhibitory threshold, so the model uses the upper
tail of the threshold distribution:

$$\theta_I = (\log[I_b(g)] - \log[I])\, t_g$$

$$g = 1 - \Phi\left(\frac{\log[I] + \theta_I / t_g - \log I_b(50)}{\sigma}\right)$$

`inhibitor_data` has tomato seeds germinated with ABA:

```{r}
fit_i <- fit_inhibitor(inhibitor_data, dose_transform = "log10")
fit_i
autoplot(fit_i)
autoplot(fit_i, type = "normalized")
```

Half of the seeds are prevented from germinating at ABA doses above
`r signif(10^coef(fit_i)[["i_b50"]], 3)` dose units.

## Predicting dose responses

With the fitted model, final germination and germination timing can be
predicted for untested doses. Pass the untransformed dose; the fit applies its
own transform:

```{r}
new <- expand.grid(GermInhibitorDosage = c(0.5, 2, 5), CumTime = c(48, 96, 192))
cbind(new, CumFraction = round(predict(fit_i, new), 3))
```

## References

Ni, B.-R., Bradford, K.J. (1992) Quantitative models characterizing seed
germination responses to abscisic acid and osmoticum. *Plant Physiol.* 98:
1057–1068. <https://doi.org/10.1104/pp.98.3.1057>

Ni, B.-R., Bradford, K.J. (1993) Germination and dormancy of abscisic acid- and
gibberellin-deficient mutant tomato seeds: sensitivity of germination to
abscisic acid, gibberellin, and water potential. *Plant Physiol.* 101:
607–617. <https://doi.org/10.1104/pp.101.2.607>
