---
title: "Seed subpopulations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Seed subpopulations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  out.width = "100%",
  dpi = 96
)
```

```{r setup}
library(pbtm)
```

The threshold models assume that one normal distribution of thresholds
describes the whole seed lot. That is often not true. Commercial lots are
frequently blends of different productions, and conditions during seed
development can leave seeds with different dormancy or vigor. A lot like this
can be modeled as a mixture of distinct subpopulations, each with its own
threshold distribution. Their germination, weighted by each subpopulation's
share of the lot, adds up to the observed time course (Bello and Bradford,
2016; Bradford and Bello, 2022). For two subpopulations:

$$g = w_1\, \Phi\left(\frac{X - \theta_{X1}/t_g - X_{b1}(50)}{\sigma_1}\right) +
(1 - w_1)\, \Phi\left(\frac{X - \theta_{X2}/t_g - X_{b2}(50)}{\sigma_2}\right)$$

where $X$ is the factor (temperature, water potential, ...) and $w_1$ is the
first subpopulation's share. Every cumulative model in pbtm (thermal time,
hydrotime, hydrothermal time, aging, promoter, and inhibitor) can be fit as a
mixture with the `subpops` argument.

## A two-subpopulation seed lot

`thermal_time_subpop_data` comes from a lot that mixes two subpopulations. A
single thermal time distribution fits it reasonably well on the usual scale:

```{r}
single <- fit_thermal_time(thermal_time_subpop_data)
single
autoplot(single)
```

The linearized scale shows the problem. A single log-normal distribution would
give a straight line at each temperature, but the data bend: the fast and slow
subpopulations follow different lines.

```{r}
autoplot(single, x_scale = "log", y_scale = "probit")
```

Fitting two subpopulations captures the bend:

```{r}
mix <- fit_thermal_time(thermal_time_subpop_data, subpops = 2)
mix
autoplot(mix, x_scale = "log", y_scale = "probit")
```

`mix$components` has one row per subpopulation with its share (`weight`) and
parameters. The flat coefficients follow the pattern `t_b1`, `t_b2`, ..., with
the free mixing fractions named `w1`, `w2`, ...

```{r}
mix$components
```

## Choosing the number of subpopulations

Adding subpopulations always improves the fit a little, so compare fits with a
criterion that penalizes the extra parameters. `subpops = "auto"` fits one to
`max_subpops` subpopulations and keeps the one with the lowest AIC:

```{r}
auto <- fit_thermal_time(thermal_time_subpop_data, subpops = "auto")
auto$subpop_table
```

Here two subpopulations are clearly better than one (a difference in AIC of
more than about 10 is strong evidence). A fit that fails to converge, as three
subpopulations often do on data like these, is left out of the table.

## Interpret with care

Mixture fits are often *equifinal*: quite different sets of component
parameters can reproduce the same time course almost equally well. With only
germination time courses to go on, a mixture can reliably show that a lot is
*not* a single population and can improve predictions. The individual
components' parameters, and even the exact number of subpopulations, are much
less certain. Some ways to firm them up:

* Collect time courses at more factor levels, which constrains the components
  much more than extra observations at the same levels.
* Hold parameters that should be shared by all components at known values.
  For example, estimate the base temperature from a single-population fit on a
  uniform lot.
* Check that the result is stable: a different `restarts` value or a
  different subset of the treatments should give similar components.
* Look for independent evidence of the subpopulations (seed size, source,
  dormancy tests).

Mixture fits start from several perturbed starting values (`restarts`, 12 by
default) and keep the best; these runs use a fixed internal random seed, so
results are reproducible and do not affect your own random number stream.

## References

Bello, P., Bradford, K.J. (2016) Single-seed oxygen consumption measurements
and population-based threshold models link respiration and germination rates
under diverse conditions. *Seed Sci. Res.* 26: 199–221.
<https://doi.org/10.1017/S0960258516000179>

Bradford, K.J., Bello, P. (2022) Applying population-based threshold models to
quantify and improve seed quality attributes. In J Buitink, O Leprince, eds,
*Advances in Seed Science and Technology for More Sustainable Crop Production*.
Burleigh Dodds Science Publishing, Cambridge, UK.
<https://doi.org/10.19103/AS.2022.0105.05>
