---
title: "Thermal time"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Thermal time}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  out.width = "100%",
  dpi = 96
)
```

```{r setup}
library(pbtm)
```

## The model

Below the optimum temperature, seeds germinate faster the further the
temperature `T` exceeds a base temperature `T_b`, below which germination does
not occur. The *thermal time* to germination of fraction `g` of the population
is constant:

$$\theta_T(g) = (T - T_b)\, t_g$$

where $t_g$ is the time to germination of fraction `g`. Thermal times vary
among seeds, following a log-normal distribution with median $\theta_T(50)$
and standard deviation $\sigma$ (on the log10 scale). The fitted model is

$$g = \Phi\left(\frac{\log_{10}[(T - T_b)\, t_g] - \log_{10}\theta_T(50)}{\sigma}\right)$$

where $\Phi$ is the standard normal cumulative distribution. The model is fit
across all temperatures at once, so only temperatures at or below the optimum
should be included.

## Data

`thermal_time_data` has tomato seeds germinated in water at three temperatures:

```{r}
head(thermal_time_data)
plot_germ_data(thermal_time_data, color = "GermTemp")
```

If your data include other treatments (for example several water potentials),
filter to a single level of those first, e.g. with `subset(data, GermWP == 0)`.
Otherwise their effects will be attributed to temperature.

## Fitting

```{r}
fit <- fit_thermal_time(thermal_time_data)
fit
summary(fit)
```

The estimated base temperature is `r round(coef(fit)[["t_b"]], 2)` °C, and half
of the seeds need `r round(coef(fit)[["theta_t50"]])` degree-hours above it to
germinate.

```{r}
autoplot(fit)
```

## Linearized plots

Because thermal time is log-normally distributed, the fitted time course at each
temperature is a straight line on a log time axis with a probit (normal
quantile) fraction axis. These lines are parallel, since every temperature
shares the same $\sigma$. Points that curve away from their line show where the
model fits poorly.

```{r}
autoplot(fit, x_scale = "log", y_scale = "probit")
```

The normalized plot converts every observation to thermal time,
$(T - T_b)\,t$, using the estimated base temperature. If the model holds, all
temperatures collapse onto one line, the population's thermal time
distribution, and the dashed line marks $\theta_T(50)$:

```{r}
autoplot(fit, type = "normalized")
```

## Using the model

The median time to germination at any sub-optimal temperature follows from the
parameters as $t_{50} = \theta_T(50) / (T - T_b)$:

```{r}
p <- coef(fit)
temps <- c(15, 20, 25)
data.frame(GermTemp = temps, predicted_t50 = p[["theta_t50"]] / (temps - p[["t_b"]]))
germ_speed(thermal_time_data, 0.5, groups = "GermTemp")
```

`predict()` gives the full predicted time course for new conditions:

```{r}
new <- expand.grid(GermTemp = 12, CumTime = c(100, 150, 200, 250))
cbind(new, CumFraction = predict(fit, new))
```

To explore alternative parameter values, hold some of them fixed:

```{r}
fit_thermal_time(thermal_time_data, fixed = list(t_b = 5))
```

## References

Bradford, K.J., Bello, P. (2022) Applying population-based threshold models to
quantify and improve seed quality attributes. In J Buitink, O Leprince, eds,
*Advances in Seed Science and Technology for More Sustainable Crop Production*.
Burleigh Dodds Science Publishing, Cambridge, UK.
<https://doi.org/10.19103/AS.2022.0105.05>
